\textbf{Using generalized additive models to analyze biomedical non-linear longitudinal data}\\ \emph{Beyond repeated measures ANOVA and Linear Mixed Models}\\ ~\\ SUPPLEMENTARY MATERIALS: APPENDIX

### 2021-06-28 Manuscripts/Manuscript by chapters- OUP submission/01-Background.Rmd
New folder for OUP Submission (Biostatistics)

Longitudinal studies are designed to repeatedly measure a variable of interest in a group (or groups) of subjects, with the intention of observing the evolution of effect across time rather than analyzing a single time point (e.g., a cross-sectional study). Biomedical research frequently uses longitudinal studies to analyze the evolution of a "treatment" effect across multiple time points; and in such studies the subjects of analysis range from animals (mice, rats, rabbits), to human patients, cells, or blood samples, among many others. Tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression [@ritter2001] [@roth2017], and cell metabolism [@jones2018] [@skala2010] are examples of the different situations where researchers have used longitudinal designs to study some  physiological response. Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals[@greening2018], to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients [@sio2016], to monthly measurements to study a long-term response such as mouth opening following radiotherapy (RT) in neck cancer patients [@kamstra2015].

Traditionally, a “frequentist” or "classical" statistical paradigm is used in biomedical research to derive inferences from a longitudinal study. The frequentist paradigm regards probability as the limit of the expected outcome when an experiment is repeated a large number of times [@wagenmakers2008], and such view is applied to the analysis of longitudinal data by assuming a null hypothesis under a statistical model that is often an _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA). The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as _complete observations_) [@gueorguieva2004] [@schober2018].

The expected linear behavior of the response through time is a key requisite in rm-ANOVA [@pinheiro2006]. This "linearity assumption" in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference. In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies. A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings [@roblyer2011] [@skala2010] [@vishwanath2009]. These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation. Therefore, when rm-ANOVA is used to draw inference of such data the estimates are inevitably biased, because the model is only able to accommodate linear trends that fail to adequately represent the biological phenomenon of interest.

>>>>

Longitudinal studies are designed to repeatedly measure a variable of interest in a group (or groups) of subjects, with the intention of observing the evolution of effect across time rather than analyzing a single time point (e.g., a cross-sectional study). Biomedical research frequently uses longitudinal studies to analyze the evolution of a "treatment" effect across multiple time points; and in such studies the subjects of analysis range from animals (mice, rats, rabbits), to human patients, cells, or blood samples, among many others. Tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression [@ritter2001; @roth2017], and cell metabolism [@jones2018; @skala2010] are examples of the different situations where researchers have used longitudinal designs to study some  physiological response. Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals[@greening2018], to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients [@sio2016], to monthly measurements to study a long-term response such as mouth opening following radiotherapy (RT) in neck cancer patients [@kamstra2015].

Traditionally, a “frequentist” or "classical" statistical paradigm is used in biomedical research to derive inferences from a longitudinal study. The frequentist paradigm regards probability as the limit of the expected outcome when an experiment is repeated a large number of times [@wagenmakers2008], and such view is applied to the analysis of longitudinal data by assuming a null hypothesis under a statistical model that is often an _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA). The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as _complete observations_) [@gueorguieva2004; @schober2018].

The expected linear behavior of the response through time is a key requisite in rm-ANOVA [@pinheiro2006]. This "linearity assumption" in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference. In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies. A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings [@roblyer2011; @skala2010; @vishwanath2009]. These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation. Therefore, when rm-ANOVA is used to draw inference of such data the estimates are inevitably biased, because the model is only able to accommodate linear trends that fail to adequately represent the biological phenomenon of interest.
